---
pagetitle: "Risk Explorer | Maternal Health Risk"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(jsonlite)

maternal_df =
  read_csv("data/Maternal Health Risk Data Set.csv") |>
  janitor::clean_names() |>
  mutate(risk_level = factor(risk_level, levels = c("low risk", "mid risk", "high risk")))
```

<div class="eyebrow">Compare Your Own Vitals</div>

# Where do you sit in this cohort?

<p class="lead">
Enter age, blood pressure, blood glucose, and resting heart rate below, and this page compares
those values two ways. First, against the `r nrow(maternal_df)` pregnancies in the dataset: a
k-nearest-neighbors lookup standardizes your inputs the same way the cohort's vitals are
(centered on the cohort mean, scaled by the cohort SD), and the 15 closest cohort records by
Euclidean distance vote on a risk level. Second, your blood pressure reading is checked against
diagnostic thresholds from a 2022 American Heart Association scientific statement on hypertension
in pregnancy — independent of this dataset, and grounded in the published clinical literature.
Everything runs in your browser — nothing you enter is sent anywhere.
</p>

```{r prep-comparison-stats}
# slim table + standardization stats for the k-NN lookup below
# (same standardization used in report.Rmd)
comparison_vars = c("age", "systolic_bp", "diastolic_bp", "bs", "heart_rate")

cohort_slim = maternal_df |>
  select(all_of(comparison_vars), risk_level) |>
  mutate(risk_level = as.character(risk_level))

cohort_stats = maternal_df |>
  summarise(across(all_of(comparison_vars), list(mean = mean, sd = sd), .names = "{.col}__{.fn}"))

cohort_json = toJSON(cohort_slim, dataframe = "rows", auto_unbox = TRUE)
stats_json  = toJSON(as.list(cohort_stats), auto_unbox = TRUE)
```

```{r embed-data, echo = FALSE, results = "asis"}
cat('<script>\n')
cat('const COHORT = ', cohort_json, ';\n', sep = "")
cat('const COHORT_STATS = ', stats_json, ';\n', sep = "")
cat('</script>\n')
```

```{r embed-widget-html, echo = FALSE, results = "asis"}
# built via cat() instead of raw HTML in the body -- pandoc mangles
# indented, blank-line-separated <div> blocks otherwise
widget_html = '
<div id="risk-explorer-app">
<div class="info-card">
<div class="field-grid">
<div>
<label for="in-age">Age (years)</label>
<input type="number" id="in-age" value="28" min="10" max="70">
</div>
<div>
<label for="in-sbp">Systolic BP (mmHg)</label>
<input type="number" id="in-sbp" value="120" min="60" max="200">
</div>
<div>
<label for="in-dbp">Diastolic BP (mmHg)</label>
<input type="number" id="in-dbp" value="80" min="40" max="120">
<div class="hint">The lower blood pressure number.</div>
</div>
<div>
<label for="in-bs">Blood glucose (mmol/L)</label>
<input type="number" id="in-bs" value="7.5" min="4" max="25" step="0.1">
<div class="hint">Fasting or random blood sugar reading.</div>
</div>
<div>
<label for="in-hr">Resting heart rate (bpm)</label>
<input type="number" id="in-hr" value="75" min="40" max="140">
</div>
</div>
<button class="btn-teal" id="btn-compare" style="border:none;cursor:pointer;">Compare to cohort</button>
<div class="result-panel" id="result-panel">
<div class="result-head">
<span>Nearest-neighbor read:</span>
<span class="badge-risk" id="result-badge"></span>
<span id="result-votes" style="color:#7c8592;font-size:0.9rem;"></span>
</div>
<div id="clinical-note"></div>
<div id="profile-chart-wrap" style="margin-top:0.9rem;display:none;">
<div id="profile-chart"></div>
</div>
<div class="chart-caption" id="profile-chart-caption" style="display:none;">Each bar is your percentile within the cohort for that vital (0-100). The dashed line marks the cohort median -- the healthy baseline this dataset is centered on.</div>
<div class="explain-box" id="result-explain"></div>
</div>
</div>
</div>
<div class="disclaimer">
This tool is descriptive, not diagnostic. The cohort comparison above tells you which historical
records in this dataset your entered vitals most resemble; the blood-pressure category is drawn
from published clinical guideline thresholds, not a diagnosis. Neither is a medical assessment,
and neither was built or validated for clinical use. If you have concerns about a pregnancy, talk
to a clinician.
</div>
'
cat(widget_html)
```

```{r embed-widget-script, echo = FALSE, results = "asis"}
# same story as the html chunk above -- keeps pandoc from touching the JS
widget_js = '
<script>
document.addEventListener("DOMContentLoaded", function () {
  var K = 15;
  var VARS = ["age", "systolic_bp", "diastolic_bp", "bs", "heart_rate"];
  var LABELS = {
    age: "Age", systolic_bp: "Systolic BP", diastolic_bp: "Diastolic BP",
    bs: "Blood glucose", heart_rate: "Heart rate"
  };

  function standardize(varName, value) {
    var mean = COHORT_STATS[varName + "__mean"];
    var sd = COHORT_STATS[varName + "__sd"];
    return (value - mean) / sd;
  }

  function nearestNeighbors(input) {
    var distances = COHORT.map(function (row) {
      var sumSq = 0;
      VARS.forEach(function (v) {
        var dz = standardize(v, input[v]) - standardize(v, row[v]);
        sumSq += dz * dz;
      });
      return { row: row, dist: Math.sqrt(sumSq) };
    });
    distances.sort(function (a, b) { return a.dist - b.dist; });
    return distances.slice(0, K);
  }

  function majorityRisk(neighbors) {
    var counts = { "low risk": 0, "mid risk": 0, "high risk": 0 };
    neighbors.forEach(function (n) { counts[n.row.risk_level]++; });
    var winner = "low risk";
    Object.keys(counts).forEach(function (k) {
      if (counts[k] > counts[winner]) winner = k;
    });
    return { winner: winner, counts: counts };
  }

  function classifyBP(sbp, dbp) {
    var reading = sbp + "/" + dbp + " mmHg";
    if (sbp >= 160 || dbp >= 110) {
      return {
        cls: "bp-severe",
        label: "Severe range (AHA 2022 guideline context)",
        text: "A reading of " + reading + " is at or above 160/110 mmHg, the threshold at which the AHA 2022 scientific statement on hypertension in pregnancy calls for antihypertensive treatment and urgent evaluation, regardless of pregnancy type."
      };
    }
    if (sbp >= 140 || dbp >= 90) {
      return {
        cls: "bp-hypertension",
        label: "Hypertension-in-pregnancy range (AHA 2022 guideline context)",
        text: "A reading of " + reading + " meets the 140/90 mmHg or higher threshold most guidelines worldwide use to diagnose hypertension in pregnancy. US (ACOG) treatment thresholds are higher, around 160/105-110 mmHg, than most international guidelines -- one of the debated points raised in the AHA statement."
      };
    }
    if (sbp >= 130 || dbp >= 80) {
      return {
        cls: "bp-elevated",
        label: "Elevated (AHA 2022 guideline context)",
        text: "A reading of " + reading + " is below the standard 140/90 mmHg pregnancy threshold, but at or above the 130/80 mmHg threshold used for the general, non-pregnant population. Reclassifying pregnancy BP readings against that lower threshold has been shown to better flag women at risk for preeclampsia (Garovic et al., 2022)."
      };
    }
    return {
      cls: "bp-normal",
      label: "Normal range (AHA 2022 guideline context)",
      text: "A reading of " + reading + " is below every BP threshold discussed in current hypertension-in-pregnancy guidelines."
    };
  }

  function percentile(varName, value) {
    var all = COHORT.map(function (r) { return r[varName]; }).sort(function (a, b) { return a - b; });
    var below = 0;
    all.forEach(function (v) { if (v <= value) below++; });
    return Math.round((below / all.length) * 100);
  }

  var SVG_NS = "http://www.w3.org/2000/svg";

  function svgEl(tag, attrs) {
    var el = document.createElementNS(SVG_NS, tag);
    for (var key in attrs) {
      if (Object.prototype.hasOwnProperty.call(attrs, key)) {
        el.setAttribute(key, attrs[key]);
      }
    }
    return el;
  }

  function makeProfileChart(input) {
    var host = document.getElementById("profile-chart");
    if (!host) return false;

    var order = ["age", "systolic_bp", "diastolic_bp", "bs", "heart_rate"];
    var pcts = order.map(function (v) { return percentile(v, input[v]); });

    var labelWidth = 108;
    var trackWidth = 190;
    var rowHeight = 40;
    var startY = 16;
    var width = labelWidth + trackWidth + 42;
    var height = startY + order.length * rowHeight + 8;

    var svg = svgEl("svg", {
      viewBox: "0 0 " + width + " " + height,
      width: "100%",
      height: String(height)
    });

    order.forEach(function (v, i) {
      var y = startY + i * rowHeight;
      var pct = pcts[i];
      var barWidth = Math.max(2, trackWidth * (pct / 100));
      var barColor = Math.abs(pct - 50) >= 25 ? "#b3323b" : "#0e7c7b";

      var labelText = svgEl("text", { x: "0", y: String(y + 13), "font-size": "11", fill: "#1c2430" });
      labelText.textContent = LABELS[v];
      svg.appendChild(labelText);

      var track = svgEl("rect", {
        x: String(labelWidth), y: String(y), width: String(trackWidth),
        height: "18", rx: "4", fill: "#eef0f5"
      });
      svg.appendChild(track);

      var bar = svgEl("rect", {
        x: String(labelWidth), y: String(y), width: String(barWidth),
        height: "18", rx: "4", fill: barColor
      });
      svg.appendChild(bar);

      var valueText = svgEl("text", {
        x: String(labelWidth + trackWidth + 8), y: String(y + 13),
        "font-size": "10", fill: "#4b5768"
      });
      valueText.textContent = pct + "th";
      svg.appendChild(valueText);
    });

    var lineX = labelWidth + trackWidth * 0.5;
    var medianLine = svgEl("line", {
      x1: String(lineX), y1: String(startY - 4),
      x2: String(lineX), y2: String(startY + order.length * rowHeight - 24),
      stroke: "#4b5768", "stroke-width": "1.5", "stroke-dasharray": "4,4"
    });
    svg.appendChild(medianLine);

    host.innerHTML = "";
    host.appendChild(svg);
    return true;
  }

  function renderResult() {
    var input = {
      age: parseFloat(document.getElementById("in-age").value),
      systolic_bp: parseFloat(document.getElementById("in-sbp").value),
      diastolic_bp: parseFloat(document.getElementById("in-dbp").value),
      bs: parseFloat(document.getElementById("in-bs").value),
      heart_rate: parseFloat(document.getElementById("in-hr").value)
    };

    var missing = VARS.filter(function (v) { return isNaN(input[v]); });
    if (missing.length) {
      alert("Please fill in all five fields before comparing.");
      return;
    }

    var neighbors = nearestNeighbors(input);
    var result = majorityRisk(neighbors);

    var badge = document.getElementById("result-badge");
    badge.textContent = result.winner;
    badge.className = "badge-risk " + (
      result.winner === "low risk" ? "badge-low" : result.winner === "mid risk" ? "badge-mid" : "badge-high"
    );

    document.getElementById("result-votes").textContent =
      "of your " + K + " nearest matches: " + result.counts["low risk"] + " low, " +
      result.counts["mid risk"] + " mid, " + result.counts["high risk"] + " high";

    var bpInfo = classifyBP(input.systolic_bp, input.diastolic_bp);
    var noteEl = document.getElementById("clinical-note");
    noteEl.innerHTML = "";
    noteEl.style.display = "block";
    noteEl.style.background = "#fbeef8";
    noteEl.style.border = "1px solid #eccbe4";
    noteEl.style.borderLeft = "4px solid #c65fae";
    noteEl.style.borderRadius = "8px";
    noteEl.style.padding = "0.85rem 1rem";
    noteEl.style.marginTop = "0.9rem";
    noteEl.style.fontSize = "0.92rem";
    noteEl.style.lineHeight = "1.55";
    noteEl.style.color = "#4b5768";

    var bpLabelEl = document.createElement("span");
    bpLabelEl.textContent = bpInfo.label;
    bpLabelEl.style.display = "block";
    bpLabelEl.style.fontSize = "0.76rem";
    bpLabelEl.style.fontWeight = "700";
    bpLabelEl.style.textTransform = "uppercase";
    bpLabelEl.style.letterSpacing = "0.04em";
    bpLabelEl.style.marginBottom = "0.3rem";
    bpLabelEl.style.color = "#93297d";
    noteEl.appendChild(bpLabelEl);
    noteEl.appendChild(document.createTextNode(bpInfo.text));

    var pctText = VARS.map(function (v) {
      return LABELS[v] + " is at the " + percentile(v, input[v]) + "th percentile";
    }).join("; ");

    document.getElementById("result-explain").innerHTML =
      "Compared against all " + COHORT.length + " records: " + pctText + " of this cohort. " +
      "The nearest-neighbor vote leans <strong>" + result.winner + "</strong>, based on the " +
      K + " cohort records closest to your inputs after standardizing each vital by the cohort mean and SD.";

    document.getElementById("result-panel").classList.add("show");

    var chartWrap = document.getElementById("profile-chart-wrap");
    var chartCaption = document.getElementById("profile-chart-caption");
    var chartOk = false;
    try {
      chartOk = makeProfileChart(input);
    } catch (err) {
      console.warn("Risk explorer: profile chart rendering failed, showing text results only.", err);
      chartOk = false;
    }
    if (chartWrap) chartWrap.style.display = chartOk ? "block" : "none";
    if (chartCaption) chartCaption.style.display = chartOk ? "block" : "none";
  }

  var btn = document.getElementById("btn-compare");
  if (btn) btn.addEventListener("click", renderResult);
});
</script>
'
cat(widget_js)
```

<footer class="site-footer">
Standardization and neighbor lookup mirror the modeling approach in the <a href="report.html">report</a>.
See the <a href="index.html">home page</a> for variable definitions and the data source.
<br><br>
Blood-pressure category thresholds cited above are from Garovic VD, Dechend R, Easterling T, et al.
<em>Hypertension in Pregnancy: Diagnosis, Blood Pressure Goals, and Pharmacotherapy: A Scientific
Statement From the American Heart Association.</em> Hypertension. 2022;79(2):e21&ndash;e41.
<a href="https://doi.org/10.1161/HYP.0000000000000208" target="_blank" rel="noopener">doi:10.1161/HYP.0000000000000208</a>.
</footer>