---
pagetitle: "Maternal Health Risk"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")

library(tidyverse)

maternal_df =
  read_csv("data/Maternal Health Risk Data Set.csv") |>
  janitor::clean_names() |>
  mutate(risk_level = factor(risk_level, levels = c("low risk", "mid risk", "high risk")))
```

<div class="eyebrow">IoT-Monitored Clinical Data &middot; Kaggle</div>

# Reading risk before it becomes an emergency

<p class="lead">
This site works through a dataset of `r nrow(maternal_df)` pregnancies monitored across hospitals,
community clinics, and maternal health cares using low-cost IoT risk sensors. Six signals — age,
systolic and diastolic blood pressure, blood glucose, body temperature, and heart rate — are used
to assign each pregnancy a clinician-labeled risk level: low, mid, or high. The goal here is the
same one behind most early-warning systems in maternal health: figure out which combinations of
ordinary vitals are already telling us something, well before an outcome forces the question.
</p>

<div class="card-row">
<a href="report.html" class="btn-magenta">Read the report &rarr;</a>
<a href="explore.html" class="btn-teal" style="background:#fff;color:#0a5b5a !important;border:1px solid #cfe8e7;">Try the risk explorer</a>
</div>

```{r stat-cards, echo = FALSE, results = "asis"}
n_total  = nrow(maternal_df)
n_high   = sum(maternal_df$risk_level == "high risk")
pct_high = round(100 * n_high / n_total, 1)

cat(
  '<div class="stat-grid">',
  '<div class="stat-card"><div class="num">', n_total, '</div><div class="label">pregnancies in the cohort</div></div>',
  '<div class="stat-card"><div class="num">6</div><div class="label">vitals recorded per record</div></div>',
  '<div class="stat-card"><div class="num">3</div><div class="label">risk levels — low, mid, high</div></div>',
  '<div class="stat-card"><div class="num">', pct_high, '%</div><div class="label">of records labeled high risk</div></div>',
  '</div>',
  sep = "\n"
)
```

## What's in the data

Each row is one pregnancy, described by six variables collected through IoT-based patient
monitoring at the point of care, plus a risk level assigned from those readings.

```{r variable-table, echo = FALSE}
tibble::tribble(
  ~Variable,      ~Description,                                     ~Units,
  "age",          "Age of the woman at the time of pregnancy",       "years",
  "systolic_bp",  "Upper value of blood pressure",                   "mmHg",
  "diastolic_bp", "Lower value of blood pressure",                   "mmHg",
  "bs",           "Blood glucose level",                             "mmol/L",
  "body_temp",    "Body temperature",                                "°F",
  "heart_rate",   "Resting heart rate",                              "bpm",
  "risk_level",   "Clinician-assigned risk during pregnancy",        "low / mid / high"
) |>
  knitr::kable()
```

Source: [Maternal Health Risk Data Set, Kaggle](https://www.kaggle.com/datasets/csafrit2/maternal-health-risk-data).
Collected from hospitals, community clinics, and maternal health cares via IoT-based risk monitoring.

## Where to go from here

<div class="card-row">
<div class="info-card">
<h3>Report</h3>
<p>Distributions, group comparisons across risk levels, and correlation structure among the six
vitals — with the analysis code available inline (click <em>Code</em>) if you want to see exactly
how a figure or number was produced.</p>
<a href="report.html">Open the report &rarr;</a>
</div>
<div class="info-card">
<h3>Risk Explorer</h3>
<p>Enter your own vitals — or someone else's — and see where they land relative to this cohort,
plus a nearest-neighbor read on which risk band they resemble most.</p>
<a href="explore.html">Open the explorer &rarr;</a>
</div>
</div>

<footer class="site-footer">
Built from the Kaggle Maternal Health Risk Data Set for exploratory and educational purposes — not a diagnostic tool.
</footer>