---
pagetitle: "Report | Maternal Health Risk"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7.5, fig.height = 4.5
)

library(tidyverse)
library(plotly)
library(nnet)
library(broom)

theme_set(
  theme_minimal(base_family = "sans") +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 12),
      strip.text = element_text(face = "bold")
    )
)

risk_colors = c("low risk" = "#1c8a5f", "mid risk" = "#b7791f", "high risk" = "#b3323b")

maternal_df =
  read_csv("data/Maternal Health Risk Data Set.csv") |>
  janitor::clean_names() |>
  mutate(risk_level = factor(risk_level, levels = c("low risk", "mid risk", "high risk")))
```

<div class="eyebrow">Exploratory Analysis</div>

# Vitals, risk levels, and where they disagree

This report works through the six IoT-monitored vitals in the maternal health risk cohort: age,
systolic and diastolic blood pressure, blood glucose (BS), body temperature, and heart rate,
against the clinician-assigned risk level (low / mid / high). Code for every table and figure
below is folded by default; click **Code** to expand it.

## Cohort characteristics

Before splitting anything by risk level, we look at the cohort as a whole: the same six vitals,
not yet grouped.

```{r table1-continuous}
cont_vars = c("age", "systolic_bp", "diastolic_bp", "bs", "body_temp", "heart_rate")
cont_labels = c(
  age = "Age (years)",
  systolic_bp = "Systolic BP (mmHg)",
  diastolic_bp = "Diastolic BP (mmHg)",
  bs = "Blood glucose (mmol/L)",
  body_temp = "Body temperature (°F)",
  heart_rate = "Heart rate (bpm)"
)

purrr::map_dfr(cont_vars, function(v) {
  x = maternal_df[[v]]
  tibble::tibble(
    Characteristic = cont_labels[[v]],
    `Mean (SD)` = paste0(round(mean(x), 2), " (", round(sd(x), 2), ")"),
    `Median [range]` = paste0(round(median(x), 2), " [", min(x), "-", max(x), "]")
  )
}) |>
  knitr::kable()
```

```{r table1-risk}
maternal_df |>
  count(risk_level) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  transmute(`Risk level` = risk_level, `n (%)` = paste0(n, " (", pct, "%)")) |>
  knitr::kable()
```

We have `r nrow(maternal_df)` records total, an average age of
`r round(mean(maternal_df$age),1)` years (SD `r round(sd(maternal_df$age),1)`), and a roughly
three-way split across risk levels rather than a dataset dominated by one category. This table
treats every row as one observation. Whether that is actually true is the subject of the next
section.

## Data quality flags

A few things in this dataset are worth flagging before we trust any number above too far. We
have not removed any of these rows in this report or in the risk explorer. The point here is to
name what is odd and be upfront that we do not know, from the data alone, whether it reflects
genuine repeat measurements, coincidence, data-entry duplication, or sensor error.

```{r dup-check}
dup_groups = maternal_df |>
  count(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate, risk_level, name = "n_copies") |>
  filter(n_copies > 1) |>
  arrange(desc(n_copies))

n_unique       = nrow(distinct(maternal_df))
n_total        = nrow(maternal_df)
n_dup_groups   = nrow(dup_groups)
n_extra_copies = sum(dup_groups$n_copies - 1)
```

```{r dup-table}
dup_groups |>
  slice_head(n = 8) |>
  knitr::kable(col.names = c("Age", "Systolic BP", "Diastolic BP", "BS", "Body Temp", "Heart Rate", "Risk Level", "Copies in dataset"))
```

Only `r n_unique` of the `r n_total` records are unique row for row. The other
`r n_extra_copies` are exact repeats: same age, same blood pressure, same glucose, same heart
rate, same risk label, of a combination that already appears elsewhere, spread across
`r n_dup_groups` distinct value combinations that each show up more than once (the most repeated
single combination appears `r max(dup_groups$n_copies)` times). That is over half the dataset.
We are leaving every row in, because deleting them assumes they are errors, and they might not
be. IoT sensors reading the same patient repeatedly, or several patients genuinely sharing
similar vitals, would produce the same pattern. It does mean every mean, SD, and correlation in
this report implicitly treats a handful of repeated value combinations as if they were
independent patients. This is worth keeping in mind when reading the risk-level split above: the
proportions shift noticeably, low risk goes from a plurality to closer to half the cohort, once
we collapse to unique rows only.

```{r age-check}
n_age_under15 = sum(maternal_df$age < 15)
n_age_under18 = sum(maternal_df$age < 18)
n_age_over55  = sum(maternal_df$age > 55)
n_hr_under40  = sum(maternal_df$heart_rate < 40)
```

Age ranges from `r min(maternal_df$age)` to `r max(maternal_df$age)` years.
`r n_age_under15` records list an age under 15, and `r n_age_under18` are under 18, roughly
`r round(100*n_age_under18/n_total,1)`% of the cohort. This dataset was collected across
hospitals and community clinics in a setting where adolescent pregnancy is a documented public
health pattern, so we are not treating the under-18 group as an error. If anything, it is a
subgroup worth its own look rather than folding into one continuous age trend. The upper end is
harder to explain the same way. `r n_age_over55` records list an age past 55, including some
close to `r max(maternal_df$age)`, past the point pregnancy is possible without medical
intervention. We treat that as a more likely candidate for a recording error, but again flag it
rather than drop it.

```{r hr-outlier}
maternal_df |>
  filter(heart_rate < 40) |>
  select(age, systolic_bp, diastolic_bp, bs, heart_rate, risk_level)
```

`r n_hr_under40` record(s) list a resting heart rate under 40 bpm, including one reading in the
single digits, not a physiologically plausible resting heart rate for a living person, let alone
a pregnant one. This is most likely a sensor dropout or transcription artifact. As with
everything else in this section, we name it rather than silently correct it.

## Cohort overview

```{r cohort-overview}
maternal_df |>
  group_by(risk_level) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  transmute(`Risk level` = risk_level, N = n, `%` = pct) |>
  knitr::kable()
```

```{r risk-bar}
p1 = maternal_df |>
  count(risk_level) |>
  ggplot(aes(x = risk_level, y = n, fill = risk_level, text = paste0(n, " records"))) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = risk_colors, guide = "none") +
  labs(x = NULL, y = "Records", title = "Distribution of risk level across the cohort")

ggplotly(p1, tooltip = "text")
```

Low-risk pregnancies are the largest single group, but the cohort is not far off a three-way
split: mid- and high-risk pregnancies together outnumber low-risk ones. That balance matters
downstream. A classifier that simply guessed "low risk" every time would still be wrong more
than half the time, so there is real separation to look for in the vitals themselves.

## Vitals by risk level

```{r summary-table}
maternal_df |>
  pivot_longer(c(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate),
               names_to = "variable", values_to = "value") |>
  group_by(variable, risk_level) |>
  summarise(cell = paste0(round(mean(value), 2), " (", round(sd(value), 2), ")"), .groups = "drop") |>
  pivot_wider(names_from = risk_level, values_from = cell) |>
  mutate(variable = cont_labels[variable]) |>
  rename(Characteristic = variable) |>
  knitr::kable(align = c("l", "r", "r", "r"))
```

Values above are mean (SD) within each risk level.

```{r boxplots, fig.height = 7}
p2 = maternal_df |>
  pivot_longer(c(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate),
               names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
    age = "Age (years)", systolic_bp = "Systolic BP (mmHg)",
    diastolic_bp = "Diastolic BP (mmHg)", bs = "Blood glucose (mmol/L)",
    body_temp = "Body temp (°F)", heart_rate = "Heart rate (bpm)")) |>
  ggplot(aes(x = risk_level, y = value, fill = risk_level)) +
  geom_boxplot(outlier.size = 0.6, width = 0.6) +
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = risk_colors, guide = "none") +
  labs(x = NULL, y = NULL, title = "Vital sign distributions across risk levels")

ggplotly(p2, height = 520) |> layout(margin = list(t = 60))
```

```{r group-stats, include = FALSE}
gs = function(var) {
  maternal_df |>
    group_by(risk_level) |>
    summarise(m = mean({{ var }}), .groups = "drop") |>
    deframe()
}
age_g   = gs(age)
sbp_g   = gs(systolic_bp)
dbp_g   = gs(diastolic_bp)
bs_g    = gs(bs)
```

The separation is clearest for blood glucose and blood pressure. High-risk pregnancies carry a
mean BS of `r round(bs_g["high risk"], 1)` mmol/L against `r round(bs_g["low risk"], 1)` mmol/L
for low-risk pregnancies, a gap that swamps the within-group spread for the low-risk group (SD
well under 1 mmol/L there, per the table above). Systolic and diastolic blood pressure move
together in the same direction: `r round(sbp_g["high risk"])`/`r round(dbp_g["high risk"])` mmHg
on average for high-risk pregnancies versus `r round(sbp_g["low risk"])`/`r round(dbp_g["low risk"])`
mmHg for low-risk ones. Age shows a real but noisier gradient. High-risk pregnancies skew roughly
`r round(age_g["high risk"] - age_g["low risk"])` years older on average than low-risk ones,
with substantial overlap in both directions. Body temperature and heart rate barely move across
groups, which is consistent with them acting more as flags for acute events (fever, tachycardia)
than as steady discriminators of risk class in this cohort.

## Correlation structure

```{r corr-heatmap}
corr_mat = maternal_df |>
  select(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate) |>
  rename(Age = age, `Systolic BP` = systolic_bp, `Diastolic BP` = diastolic_bp,
         BS = bs, `Body Temp` = body_temp, `Heart Rate` = heart_rate) |>
  cor() |>
  round(2)

corr_long = corr_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r")

p3 = corr_long |>
  ggplot(aes(x = var1, y = var2, fill = r, text = paste0(var1, " × ", var2, ": ", r))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#0e7c7b", mid = "#f6f7fb", high = "#b3323b", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "Pairwise correlation among the six vitals", fill = "r") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

ggplotly(p3, tooltip = "text")
```

Systolic and diastolic blood pressure are, unsurprisingly, the strongest pair in the matrix
(r = `r round(corr_mat["Systolic BP", "Diastolic BP"], 2)`), since they measure the same
physiological event from two ends. Age correlates moderately with blood pressure and glucose
alike (r in the `r round(min(corr_mat["Age", c("Systolic BP","Diastolic BP","BS")]), 2)` to
`r round(max(corr_mat["Age", c("Systolic BP","Diastolic BP","BS")]), 2)` range), which tracks
with the age gradient seen in the boxplots above. Heart rate stands apart. Its strongest
correlation with anything else in the matrix is
`r round(max(abs(corr_mat["Heart Rate", c("Age","Systolic BP","Diastolic BP","BS","Body Temp")])), 2)`,
weak enough that it is carrying mostly independent information, or mostly noise, relative to the
other five vitals.

## A simple risk classifier

As a baseline read on how separable the three risk levels actually are from the six vitals, we
fit a multinomial logistic regression on `r round(0.75 * nrow(maternal_df))` randomly sampled
records and evaluated it on the held-out `r nrow(maternal_df) - round(0.75 * nrow(maternal_df))`.
This is the same modeling approach behind the risk explorer's comparison logic, formalized here
with a train and test split. We standardize each vital to a common scale (mean 0, SD 1, using
the training set's own mean and SD) before fitting, so the coefficients below are comparable to
each other regardless of the units the underlying vital happens to be measured in.

```{r classifier}
set.seed(2026)

split_idx = sample(seq_len(nrow(maternal_df)), size = round(0.75 * nrow(maternal_df)))
train_df  = maternal_df[split_idx, ]
test_df   = maternal_df[-split_idx, ]

scale_vars  = c("age", "systolic_bp", "diastolic_bp", "bs", "body_temp", "heart_rate")
train_means = sapply(train_df[scale_vars], mean)
train_sds   = sapply(train_df[scale_vars], sd)

standardize_df = function(df) {
  for (v in scale_vars) df[[v]] = (df[[v]] - train_means[[v]]) / train_sds[[v]]
  df
}

train_std = standardize_df(train_df)
test_std  = standardize_df(test_df)

fit = multinom(
  risk_level ~ age + systolic_bp + diastolic_bp + bs + body_temp + heart_rate,
  data = train_std, trace = FALSE
)

pred = predict(fit, newdata = test_std)
acc  = mean(pred == test_df$risk_level)
```

```{r confusion-heatmap}
conf_df = tibble(Predicted = pred, Actual = test_df$risk_level) |>
  count(Predicted, Actual, name = "n") |>
  complete(Predicted = levels(test_df$risk_level), Actual = levels(test_df$risk_level), fill = list(n = 0)) |>
  mutate(label_color = if_else(n > max(n) * 0.55, "white", "#1c2430"))

p4 = conf_df |>
  ggplot(aes(x = Actual, y = Predicted, fill = n,
             text = paste0("Predicted ", Predicted, ", actual ", Actual, ": ", n))) +
  geom_tile(color = "#f6f7fb", linewidth = 1.2) +
  geom_text(aes(label = n, color = label_color), fontface = "bold", size = 4) +
  scale_color_identity() +
  scale_fill_gradient(low = "#fbeef8", high = "#7a1f6e") +
  labs(x = "Actual", y = "Predicted", title = "Confusion matrix: predicted vs. actual risk level", fill = "Count")

ggplotly(p4, tooltip = "text")
```

Held-out accuracy is **`r scales::percent(acc, accuracy = 0.1)`**, against a
`r scales::percent(max(prop.table(table(test_df$risk_level))), accuracy = 0.1)` baseline for
always predicting the majority class in the test fold. The darkest cells above sit on the
diagonal, which is where predicted and actual risk level agree; the confusion matrix shows where
the model actually struggles. Mid-risk pregnancies are the hardest to separate cleanly, which
lines up with the boxplots: mid-risk vitals sit between the low- and high-risk groups on almost
every variable, rather than forming a distinct cluster of their own.

```{r term-labels, include = FALSE}
var_label_map = c(
  age = "age", systolic_bp = "systolic blood pressure", diastolic_bp = "diastolic blood pressure",
  bs = "blood glucose", body_temp = "body temperature", heart_rate = "heart rate"
)
```

The model is fit with low risk as the reference category, since it is the first level of
`risk_level`. Every coefficient below is therefore relative to a low-risk pregnancy: a positive
estimate means that vital, holding the others fixed, pulls the classification away from low risk
and toward mid or high risk, and a negative one pulls back toward low risk. Rows are grouped by
vital, high risk above mid risk, so the two effects for a given vital sit next to each other.

```{r coef-table}
tidy(fit) |>
  filter(term != "(Intercept)") |>
  mutate(
    across(where(is.numeric), ~round(.x, 3)),
    term = str_to_sentence(var_label_map[term])
  ) |>
  arrange(term, y.level) |>
  transmute(
    `Risk level` = y.level, Variable = term, Estimate = estimate,
    `Std. Error` = std.error, z = statistic, p = p.value
  ) |>
  knitr::kable(table.attr = 'class="coef-table"')
```

```{r top-coef, include = FALSE}
coef_rank = tidy(fit) |>
  filter(y.level == "high risk", term != "(Intercept)") |>
  arrange(desc(abs(estimate)))

top1_label = var_label_map[[coef_rank$term[1]]]
top2_label = var_label_map[[coef_rank$term[2]]]
```

`r str_to_sentence(top1_label)` carries the largest standardized pull toward a high-risk
classification in this model, followed by `r top2_label`. That lines up with the group means
above and with what is clinically expected: hyperglycemia and hypertensive markers are the two
vitals most directly tied to pregnancy risk in the obstetric literature this dataset is meant to
echo.

<footer class="site-footer">
Analysis run against the Kaggle Maternal Health Risk Data Set. See the <a href="index.html">home page</a>
for variable definitions, or the <a href="explore.html">risk explorer</a> to compare individual
vitals against this cohort.
</footer>