---
title: "Report"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7.5, fig.height = 4.5
)

library(tidyverse)
library(plotly)
library(nnet)
library(broom)

theme_set(
  theme_minimal(base_family = "sans") +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 12),
      strip.text = element_text(face = "bold")
    )
)

risk_colors = c("low risk" = "#1c8a5f", "mid risk" = "#b7791f", "high risk" = "#b3323b")

maternal_df =
  read_csv("data/Maternal Health Risk Data Set.csv") |>
  janitor::clean_names() |>
  mutate(risk_level = factor(risk_level, levels = c("low risk", "mid risk", "high risk")))
```

<div class="eyebrow">Exploratory Analysis</div>

# Vitals, risk levels, and where they disagree

This report works through the six IoT-monitored vitals in the maternal health risk cohort —
age, systolic and diastolic blood pressure, blood glucose (BS), body temperature, and heart
rate — against the clinician-assigned risk level (low / mid / high). Code for every table and
figure below is folded by default; click **Code** to expand it.

## Cohort overview

```{r cohort-overview}
maternal_df |>
  group_by(risk_level) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r risk-bar}
p1 = maternal_df |>
  count(risk_level) |>
  ggplot(aes(x = risk_level, y = n, fill = risk_level, text = paste0(n, " records"))) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = risk_colors, guide = "none") +
  labs(x = NULL, y = "Records", title = "Distribution of risk level across the cohort")

ggplotly(p1, tooltip = "text")
```

Low-risk pregnancies are the largest single group, but the cohort is not far off a three-way
split — mid- and high-risk pregnancies together outnumber low-risk ones. That balance matters
downstream: a classifier that simply guessed "low risk" every time would still be wrong more
than half the time, so there's real separation to look for in the vitals themselves.

## Vitals by risk level

```{r summary-table}
maternal_df |>
  pivot_longer(c(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate),
               names_to = "variable", values_to = "value") |>
  group_by(variable, risk_level) |>
  summarise(mean = round(mean(value), 2), sd = round(sd(value), 2), .groups = "drop") |>
  pivot_wider(names_from = risk_level, values_from = c(mean, sd)) |>
  knitr::kable()
```

```{r boxplots, fig.height = 7}
p2 = maternal_df |>
  pivot_longer(c(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate),
               names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
    age = "Age (years)", systolic_bp = "Systolic BP (mmHg)",
    diastolic_bp = "Diastolic BP (mmHg)", bs = "Blood glucose (mmol/L)",
    body_temp = "Body temp (°F)", heart_rate = "Heart rate (bpm)")) |>
  ggplot(aes(x = risk_level, y = value, fill = risk_level)) +
  geom_boxplot(outlier.size = 0.6, width = 0.6) +
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = risk_colors, guide = "none") +
  labs(x = NULL, y = NULL, title = "Vital sign distributions across risk levels")

ggplotly(p2, height = 520) |> layout(margin = list(t = 60))
```

```{r group-stats, include = FALSE}
gs = function(var) {
  maternal_df |>
    group_by(risk_level) |>
    summarise(m = mean({{ var }}), .groups = "drop") |>
    deframe()
}
age_g   = gs(age)
sbp_g   = gs(systolic_bp)
dbp_g   = gs(diastolic_bp)
bs_g    = gs(bs)
```

The separation is clearest for blood glucose and blood pressure. High-risk pregnancies carry a
mean BS of `r round(bs_g["high risk"], 1)` mmol/L against `r round(bs_g["low risk"], 1)` mmol/L
for low-risk pregnancies — a gap that swamps the within-group spread for the low-risk group
(SD well under 1 mmol/L there, per the table above). Systolic and diastolic blood pressure move
together in the same direction: `r round(sbp_g["high risk"])`/`r round(dbp_g["high risk"])` mmHg
on average for high-risk pregnancies versus `r round(sbp_g["low risk"])`/`r round(dbp_g["low risk"])`
mmHg for low-risk ones. Age shows a real but noisier gradient — high-risk pregnancies skew
roughly `r round(age_g["high risk"] - age_g["low risk"])` years older on average than low-risk
ones, with substantial overlap in both directions. Body temperature and heart rate barely move
across groups, which is consistent with them acting more as flags for acute events (fever,
tachycardia) than as steady discriminators of risk class in this cohort.

## Correlation structure

```{r corr-heatmap}
corr_mat = maternal_df |>
  select(age, systolic_bp, diastolic_bp, bs, body_temp, heart_rate) |>
  rename(Age = age, `Systolic BP` = systolic_bp, `Diastolic BP` = diastolic_bp,
         BS = bs, `Body Temp` = body_temp, `Heart Rate` = heart_rate) |>
  cor() |>
  round(2)

corr_long = corr_mat |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r")

p3 = corr_long |>
  ggplot(aes(x = var1, y = var2, fill = r, text = paste0(var1, " × ", var2, ": ", r))) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#0e7c7b", mid = "#f6f7fb", high = "#b3323b", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "Pairwise correlation among the six vitals", fill = "r") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

ggplotly(p3, tooltip = "text")
```

Systolic and diastolic blood pressure are, unsurprisingly, the strongest pair in the matrix
(r = `r round(corr_mat["Systolic BP", "Diastolic BP"], 2)`) — they're measuring the same
physiological event from two ends. Age correlates moderately with blood pressure and glucose
alike (r in the `r round(min(corr_mat["Age", c("Systolic BP","Diastolic BP","BS")]), 2)`–`r round(max(corr_mat["Age", c("Systolic BP","Diastolic BP","BS")]), 2)`
range), which tracks with the age gradient seen in the boxplots above. Heart rate stands apart —
its strongest correlation with anything else in the matrix is
`r round(max(abs(corr_mat["Heart Rate", c("Age","Systolic BP","Diastolic BP","BS","Body Temp")])), 2)`,
which is weak enough that it's carrying mostly independent information (or mostly noise) relative
to the other five vitals.

## A data quality note

```{r hr-outlier}
maternal_df |>
  filter(heart_rate < 40) |>
  select(age, systolic_bp, diastolic_bp, bs, heart_rate, risk_level)
```

`r sum(maternal_df$heart_rate < 40)` record(s) list a resting heart rate under 40 bpm, including
at least one reading in the single digits — not a physiologically plausible resting heart rate
for a pregnant adult, and more likely a sensor dropout or transcription artifact than a true
reading. These rows are left in the descriptive summaries above (they're a small share of the
cohort and don't move the group means meaningfully) but are worth flagging before this data feeds
anything with real stakes attached.

## A simple risk classifier

As a baseline read on how separable the three risk levels actually are from the six vitals, a
multinomial logistic regression was fit on `r round(0.75 * nrow(maternal_df))` randomly sampled
records and evaluated on the held-out `r nrow(maternal_df) - round(0.75 * nrow(maternal_df))`.
This is the same modeling approach behind the risk explorer's comparison logic, just formalized
with a train/test split here.

```{r classifier}
set.seed(2026)

split_idx = sample(seq_len(nrow(maternal_df)), size = round(0.75 * nrow(maternal_df)))
train_df  = maternal_df[split_idx, ]
test_df   = maternal_df[-split_idx, ]

fit = multinom(
  risk_level ~ age + systolic_bp + diastolic_bp + bs + body_temp + heart_rate,
  data = train_df, trace = FALSE
)

pred = predict(fit, newdata = test_df)
acc  = mean(pred == test_df$risk_level)

table(Predicted = pred, Actual = test_df$risk_level)
```

Held-out accuracy: **`r scales::percent(acc, accuracy = 0.1)`**, against a
`r scales::percent(max(prop.table(table(test_df$risk_level))), accuracy = 0.1)` baseline for
always predicting the majority class in the test fold. The confusion matrix above shows where
the model actually struggles — mid-risk pregnancies are the hardest to separate cleanly, which
lines up with the boxplots: mid-risk vitals sit between the low- and high-risk groups on almost
every variable, rather than forming a distinct cluster of their own.

```{r coef-table}
tidy(fit) |>
  filter(term != "(Intercept)") |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  arrange(y.level, term) |>
  knitr::kable()
```

Blood glucose carries the largest standardized pull toward a high-risk classification, followed
by blood pressure — consistent with the group means above and with what's clinically expected:
hyperglycemia and hypertensive markers are the two vitals most directly tied to pregnancy risk
in the obstetric literature this dataset is meant to echo.

<footer class="site-footer">
Analysis run against the Kaggle Maternal Health Risk Data Set. See the <a href="index.html">home page</a>
for variable definitions, or the <a href="explore.html">risk explorer</a> to compare your own vitals
against this cohort.
</footer>
